Dependent Types for Safe and Secure Web Programming

### 2014-02-03 paper/conclusion.tex
More paper revisions...

Type providers, as originally implemented in F\# \cite{msr:tp}, are an
interesting method by which external data sources may be used to import
external information, such that it may be used during compilation. 
In this way, it becomes possible to use the extra type
information to statically ensure the validity of artefacts such as SQL
queries and data structures. If data structures within the program do not
conform to a given database schema, for example, then the program will not
type-check.  This has been implemented for 
\idris{} 
\cite{christiansen:dtp}, exploiting the fact that types can be
calculated by functions to avoid unsafely generating extra code in the type provider
step. We believe that this technique would be an interesting avenue of exploration to provide further static guarantees about the form of database queries.

>>>>

Type providers, as originally implemented in F\# \cite{msr:tp}, allow external data sources to be used to import
external information to be used during compilation. 
In this way, it becomes possible to use the extra type
information to statically ensure the validity of artefacts such as SQL
queries and data structures. If data structures within the program do not
conform to a given database schema, for example, then the program will not
type-check.  Type providers have been implemented for 
\idris{} 
\cite{christiansen:dtp}, exploiting the fact that types can be
calculated by functions to avoid unsafely generating extra code in the type provider
step. We believe that this technique would be an interesting avenue of exploration to provide further static guarantees about the form of database queries.

### 2014-02-03 paper/forms.tex
(Hopefully) final revision of paper before IFL2013 submission.
Paper revisions
More paper revisions...

\section{Type-aware form handling}

@@

cgiHello : CGIProg [] ()
cgiHello = do
  handler_set <- isHandlerSet
  if handler_set then do
    handleForm handlers
    return ()
  else do
    addForm "nameform" "helloform" showHelloForm
    return ()

@@

\noindent
Again, we use \texttt{Effects} to build a form.
By recording
the type of each form element as it is added in the type of the form, we may
statically ensure that the user-supplied handler function is of the correct
type to handle the data supplied by the form: using an
incompatible handler will result in a compile-time type error. The \texttt{Form}
effect and associated resource \texttt{FormRes} is given in Figure 
\ref{fig:formeffect}.
The \texttt{using} notation here indicates that within
the block, where \texttt{G} and \texttt{E} occur they are implicit arguments with
the given type.

The general process of form construction is illustrated by the \texttt{AddTextBox}
and \texttt{Submit} operations of the \texttt{Form} effect:

\begin{SaveVerbatim}{formeffsmall}

data Form : Effect where
  AddTextBox : (label : String) -> (fty : FormTy) -> 
               Maybe (interpFormTy fty) -> 
               Form (FormRes G E) 
                    (FormRes (fty :: G) E) () 
     ...
  Submit : (mkHandlerFn ((reverse G), E)) -> String -> 
           Form (FormRes G E) (FormRes [] []) String

\end{SaveVerbatim}
\useverb{formeffsmall}

\noindent
These use the resource associated with the effect, \texttt{FormRes}, to
construct the form. Adding a field such as a text box adds a new type,
\texttt{fty} to the list of field types, carried in the resource. When the form
is complete, the \texttt{Submit} operation adds a submit button and returns the
HTML text for the form, flushing the list of field types, and using it to
construct the type for an appropriate handler function.
%
To specify a form instance, we define a function of type \texttt{UserForm}:

@@

UserForm : Type
UserForm = EffM m [FORM (FormRes []) 
                        (FormRes [])] String

\end{SaveVerbatim}
\useverb{userform}

@@

\begin{SaveVerbatim}{interpwebeff}

webEffect : WebEffect -> EFFECT

\end{SaveVerbatim}
\useverb{interpwebeff}

Whilst it is not possible to serialise arbitrary effects due to the associated
difficulties with serialising initial resource environments, we allow for three
effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
This is, however, not an inherent limitation as the \texttt{Effects} library
permits introduction of additional effects within an effectful computation.
%
We may specify a handler function of type \texttt{FormHandler}:

@@

\noindent
In order to associate a handler with a form, we may call the \texttt{addSubmit}
function:

@@

\noindent
This function takes a handler function and a list of available handlers, along with
an automatically constructed proof (using the a \texttt{default} argument) that the
handler is available.
Let us look at each aspect of this function in turn. Firstly, the
\texttt{mkHandlerFn} function calculates the required type of the handler
function from the list of types associated with the form elements, and the
effects we specified with \texttt{useEffects}. Note that since we prepend types
to the list of \texttt{FormTy}s as opposed to appending them, we must reverse
the list.

@@

\noindent
The \texttt{**} notation denotes a dependent pair, in which the type of the second
element of the pair is parameterised over the value of the first element. It is
an existential binding;
the notation \texttt{(x ** P x)} can be read as ``there exists an \texttt{x} such that 
\texttt{P x} holds''.
Therefore a \texttt{HandlerFn} states that there exists a function type
\texttt{ft} such that we have a handler for it, and a unique string identifier
which is used to serialise a
reference to the handler function.

@@

\noindent
\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the
form elements and the effects used by the handler function. \texttt{FnHere}
is a proof that the element is at the head of the current point of the list,
whereas \texttt{FnThere} is a proof that the element is in the tail of 
the list. 
%TODO: this should probably be rewritten
We then use an automatic proof search to
generate the proof at compile time, should one exist. The proof
may then be used in subsequent computations: we use it to retrieve
the unique identifier for the function. If the automated proof search fails,
compilation will fail.

>>>>

\section{Dependently-typed Form Handling}

@@

cgiHello : CGIProg [] ()
cgiHello = do
  handler_set <- isHandlerSet
  if handler_set then do
    handleForm handlers
    return ()
  else do
    addForm showHelloForm
    return ()

@@

\noindent
Again, we use \texttt{Effects} to build a form.
By recording
the type of each form element as it is added in the type of the form, we may
statically ensure that the user-supplied handler function is of the correct
type to handle the data supplied by the form: using an
incompatible handler will result in a compile-time type error. The \texttt{Form}
effect and associated resource \texttt{FormRes} is given in Figure 
\ref{fig:formeffect}.

The \texttt{using} notation indicates that where \texttt{G} and \texttt{E} occur within the block, they correspond to \emph{implicit arguments} within the type of the form construction resource. 
\texttt{G} is of type \texttt{List FormTy}, indicating the types of the elements within the form, and \texttt{E} is of type \texttt{List WebEffect}. This allows us to record the types of form elements, which we may later use to ensure that the type of the handler function corresponds to the types of the elements within the form.
%The \texttt{using} notation here indicates that within
%the block, where \texttt{G} and \texttt{E} occur they are implicit arguments with
%the given resource type.

The general process of form construction is illustrated by the \texttt{AddTextBox}
and \texttt{Submit} operations of the \texttt{Form} effect, shown in Figure ~\ref{fig:formeffect}. These use the resource associated with the effect, \texttt{FormRes}, to construct the form. Adding a field such as a text box adds a new type,
\texttt{fty} to the list of field types, carried in the resource. When the form
is complete, the \texttt{Submit} operation adds a submit button and returns the
HTML text for the form, flushing the list of field types, and using it to
construct the type for an appropriate handler function.
%
To specify a form instance, we define a function of type \texttt{UserForm}:

@@

UserForm : Type
UserForm = EffM m [FORM (FormRes []) 
                        (FormRes [])] String
\end{SaveVerbatim}
\useverb{userform}

@@

\begin{SaveVerbatim}{interpwebeff}
webEffect : WebEffect -> EFFECT
\end{SaveVerbatim}
\useverb{interpwebeff}

While it is not possible to serialise arbitrary effects due to the associated
difficulties with serialising initial resource environments, we allow for three
effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
This is, however, not an inherent limitation as the \texttt{Effects} library
permits introduction of additional effects within an effectful computation.

We may specify a handler function of type \texttt{FormHandler}:

@@

\noindent
In order to associate a handler with a form, we call the \texttt{addSubmit}
function:

@@

\noindent
This function takes a handler function and a list of available handlers, along with
an automatically constructed proof (using the \texttt{default} argument) that the
handler is available.
Let us look at each aspect of this function in turn. Firstly, the
\texttt{mkHandlerFn} function calculates the required type of the handler
function from the list of types associated with the form elements, and the
effects we specified with \texttt{useEffects}. Note that since we prepend types
to the list of \texttt{FormTy}s as opposed to appending them, we must reverse
the list of form elements \texttt{G}.

@@

\noindent
The \texttt{**} notation denotes a dependent pair, in which the type of the second
element of the pair is parameterised over the value of the first element. It is
an existential binding:
the notation \texttt{(x ** P x)} can be read as ``there exists an \texttt{x} such that 
\texttt{P x} holds''.
Therefore a \texttt{HandlerFn} states that there exists a function type
\texttt{ft} such that we have a handler for it, and a unique string identifier
which is used to serialise a
reference to the handler function.

@@

\noindent
\texttt{FnElem} is parameterised over \texttt{G} and \texttt{E}, the types of the
form elements and the effects used by the handler function. \texttt{FnHere}
is a proof that the element is at the head of the current point of the list,
whereas \texttt{FnThere} is a proof that the element is in the tail of 
the list. 
We then use an automatic proof search to
generate the proof at compile time, should one exist. The proof
may then be used in subsequent computations: we use it to retrieve
the unique identifier for the function. If the automated proof search fails,
compilation will fail.

### 2014-02-03 paper/introduction.tex
More paper revisions...

Web applications, whilst ubiquitous, are also prone to incorrect construction
and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
scripting \cite{owasp:xss}. Security breaches using such exploits are
far-reaching, and high profile cases involve large corporations such as Sony,
who suffered a well-publicised and extremely costly SQL Injection breach in
2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
\cite{imperva:yahoo}.

@@

%%------- EXAMPLE
Let us consider a simple database access routine, written in
PHP, where we wish to obtain the name and address of every employee working in
a given department, \texttt{\$dept}. We firstly construct an object
representing a database connection, where the arguments are the database host,
user, password and name respectively:

@@

Problems may arise if the protocol is not followed correctly.
A developer may, for example, accidentally close a statement whilst still
retrieving rows, which would cause a runtime error. Similarly, a programmer may
omit closing the statement or connection, which can lead to
problems such as resource leaks in longer-running server applications.
However, in conventional programming languages, there is no way to check
automatically, at compile-time, that a protocol is followed.

In contrast, the use of \textit{dependent types} makes it possible
to specify a program's behaviour precisely, and to check that a 
specification is followed.
%
Unfortunately, automatic verification by a compiler can be difficult or
often impossible, requiring additional proofs to be given by the programmer.

>>>>

Web applications, whilst ubiquitous, are also prone to incorrect construction
and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
scripting \cite{owasp:xss}. Security breaches are
far-reaching, and high profile cases involve large corporations such as Sony,
who suffered a well-publicised and extremely costly SQL Injection breach in
2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
\cite{imperva:yahoo}.

@@

%%------- EXAMPLE
\noindent
Let us consider a simple database access routine, written in
PHP, where we wish to obtain the name and address of every employee working in
a given department, \texttt{\$dept}. We firstly construct an object
representing a database connection, where the arguments are the database host,
user, password and name respectively:

@@

Problems may arise if the protocol is not followed correctly.
A developer may, for example, accidentally close a statement whilst still
retrieving rows, which would cause a runtime error. Similarly, a developer may
omit closing the statement or connection, which can lead to
problems such as resource leaks in longer-running server applications.
However, in conventional programming languages, there is no way to check
automatically, at compile-time, that a protocol is followed.

In contrast, the use of \textit{dependent types} makes it possible
to specify a program's behaviour precisely, and to check that a 
specification is followed.
%
Unfortunately, automatic verification by a compiler can be difficult or
often impossible, requiring additional proofs to be given by the developer.

### 2014-02-03 paper/messageboard.tex
(Hopefully) final revision of paper before IFL2013 submission.
More paper revisions...

\begin{SaveVerbatim}{handlereq}

handleRequest : CGIProg 
     [SESSION (SessionRes SessionUninitialised), 
      SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.return ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)

@@

\noindent
Once we have loaded the session data from the database, we then check whether
the \texttt{UserID} variable is set, which demonstrates that a user has
successfully logged into the system, and allows us to use the ID in subsequent
computations. The database operation to insert the thread into the database is
performed by \texttt{threadInsert}, shown in Figure \ref{fig:threadins}.
This uses a library function \texttt{executeInsert}, which abstracts over the
low-level resource usage protocol, enabling for provably-correct database
access without the excess boilerplate code. In addition, \texttt{executeInsert}
returns the unique row ID of the last item which was inserted, which may be
used in subsequent computations. In the case of the message board, we use this
to associate the first post of the thread with the thread being inserted.

@@

\begin{figure}[h]
\useverb{traversethreads}
\caption{Thread Insertion}
\label{fig:traverse}
\end{figure}

@@

\begin{SaveVerbatim}{handlelogin}
handleLoginForm (Just name) (Just pwd) = do
  auth_res <- lift' (authUser name pwd)
  case auth_res of
    Right (Just uid) => do
      set_sess_res <- setSession uid
      if set_sess_res then do
        lift' (output $ "Welcome, " ++ name)
        return ()
      else do
        lift' (output "Could not set session")
         return ()
    Right Nothing => do
      lift' (output "Invalid username or password")
      return ()
    Left err => do
      lift' (output $ "Error: " ++ err)
      return ()
\end{SaveVerbatim}

\begin{figure}[h]
\useverb{handlelogin}
\caption{Thread Insertion}
\label{fig:handlelogin}
\end{figure}

@@

Although we have described a relatively simple application, we have shown that
through the use of embedded domain-specific languages, 
and particularly by encapsulating resource usage protocols in the types,
we can write verified
code that fails to compile should resources be incorrectly accessed.
Additionally, we have used the form handling mechanism to simply handle the
arguments passed by the user. Importantly, we have shown that dependent types
can be used to increase confidence in an (albeit simplified) real-world
application, without requiring developers to supply proofs or indeed work
explicitly with dependent types.

>>>>

\begin{SaveVerbatim}{handlereq}
 handleRequest : 
   CGIProg [SESSION (SessionRes SessionUninitialised), 
            SQLITE ()] ()
 handleRequest = do 
   handler_set <- isHandlerSet
   if handler_set then do
     handleForm handlers
     return ()
   else do
     action <- queryGetVar "action"
     thread_id <- queryGetVar "thread_id"
     handleNonFormRequest action (map strToInt thread_id)

@@

Once we have loaded the session data from the database, we then check whether
the \texttt{UserID} variable is set, which demonstrates that a user has
successfully logged into the system, and allows us to use the ID in subsequent
computations. The database operation to insert the thread into the database is
performed by \texttt{threadInsert}, shown in Figure \ref{fig:threadins}.

This uses a library function \texttt{executeInsert}, which abstracts over the
low-level resource usage protocol, enabling for provably-correct database
access without the additional DSL code. In addition, \texttt{executeInsert}
returns the unique row ID of the last item which was inserted, which may be
used in subsequent computations. In the case of the message board, we use this
to associate the first post of the thread with the thread being inserted.

@@

\begin{figure}[h]
\useverb{traversethreads}
\caption{Listing Threads}
\label{fig:traverse}
\end{figure}

@@

\begin{SaveVerbatim}{handlelogin}
handleLoginForm (Just name) (Just pwd) = do
  auth_res <- (authUser name pwd)
  case auth_res of
    Right (Just uid) => do
      set_sess_res <- setSession uid
      if set_sess_res then do
        output $ "Welcome, " ++ name
        return ()
      else do
         output "Could not set session"
         return ()
    Right Nothing => do
      output "Invalid username or password"
      return ()
    Left err => do
      output $ "Error: " ++ err
      return ()
\end{SaveVerbatim}

\begin{figure}[h]
\useverb{handlelogin}
\caption{Authentication}
\label{fig:handlelogin}
\end{figure}

@@

Although we have described a relatively simple application, we have shown that
through the use of embedded domain-specific languages, 
and particularly by encapsulating resource usage protocols in the types,
we can write verified
code that fails to compile should resources be incorrectly accessed.
Additionally, we have used the form handling mechanism to simply handle the
arguments passed by the user. Importantly, we have shown that dependent types
can be used to increase confidence in an (albeit simplified) real-world
application, without requiring developers to supply proofs.
